%##########################################################################
% CHAPTER 2: LITERATURE REVIEW
%##########################################################################
\chapter{Literature Review}

\section{Chapter Overview}

This chapter surveys the six bodies of research that underpin the system design: narrative personalization, player modelling, Flow theory, large language models for game narrative, reinforcement learning for adaptation, and episodic memory. For each area, the review identifies what prior work accomplished and, more importantly, where it stopped short, building incrementally toward the research gap that this dissertation sets out to fill.

Figure~\ref{fig:litmap} maps the six domains and their inter-relationships. The solid arrows show how each domain feeds into the present work; the dashed cross-links highlight dependencies between domains that existing literature has acknowledged but not yet exploited within a single system.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    scale=0.75, every node/.style={scale=0.75},
    >={Stealth[length=2mm]},
    dbox/.style={
        rectangle, rounded corners=4pt, draw=black!55, line width=0.7pt,
        minimum width=3.4cm, minimum height=1.1cm, align=center,
        font=\small, inner sep=5pt
    },
    central/.style={
        rectangle, rounded corners=6pt, draw=black!70, line width=1.2pt,
        fill=black!8, minimum width=4.2cm, minimum height=1.3cm,
        align=center, font=\small\bfseries, inner sep=6pt
    },
    feed/.style={->, line width=0.7pt, black!55},
    cross/.style={<->, line width=0.5pt, black!30, dashed}
]

% ── Central node ──
\node[central] (thiswork) {This Work:\\Closed-Loop\\Narrative Plugin};

% ── Six domains arranged around the centre ──
\node[dbox, fill=blue!10, above left=1.8cm and 2.8cm of thiswork] (pm) {
    Player Modelling\\
    {\scriptsize Bartle, Hexad, player2vec}
};

\node[dbox, fill=green!10, above=2.2cm of thiswork] (flow) {
    Flow Theory\\
    {\scriptsize Csikszentmihalyi, Chen}
};

\node[dbox, fill=orange!10, above right=1.8cm and 2.8cm of thiswork] (rl) {
    Reinforcement Learning\\
    {\scriptsize DDA, PPO}
};

\node[dbox, fill=red!8, below left=1.8cm and 2.8cm of thiswork] (np) {
    Narrative\\Personalization\\
    {\scriptsize Fa\c{c}ade, Tracery, branching}
};

\node[dbox, fill=purple!10, below=2.2cm of thiswork] (llm) {
    LLMs for Games\\
    {\scriptsize PANGeA, LIGS, RAG}
};

\node[dbox, fill=yellow!15, below right=1.8cm and 2.8cm of thiswork] (em) {
    Episodic Memory\\
    {\scriptsize EM-LLM}
};

% ── Arrows into central node ──
\draw[feed] (pm) -- (thiswork);
\draw[feed] (flow) -- (thiswork);
\draw[feed] (rl) -- (thiswork);
\draw[feed] (np) -- (thiswork);
\draw[feed] (llm) -- (thiswork);
\draw[feed] (em) -- (thiswork);

% ── Cross-links between related domains ──
\draw[cross] (pm) -- node[above, font=\tiny\itshape, text=black!50] {state signal} (flow);
\draw[cross] (flow) -- node[above, font=\tiny\itshape, text=black!50] {reward} (rl);
\draw[cross] (np) -- node[below, font=\tiny\itshape, text=black!50] {generation} (llm);
\draw[cross] (llm) -- node[below, font=\tiny\itshape, text=black!50] {context} (em);

\end{tikzpicture}
\caption[Literature map of reviewed research domains]{Literature map of reviewed research domains.}
\label{fig:litmap}
\end{figure}

%--------------------------------------------------------------------------
\section{Background to the Problem}
%--------------------------------------------------------------------------

The question of how games might deliver individualized narrative experiences has occupied researchers for the better part of three decades, and the answers have evolved through several distinct paradigms, each solving part of the problem while leaving a conspicuous remainder.

The earliest serious attempts were structural. Drama management systems, exemplified by Mateas and Stern's Fa\c{c}ade project \citep{mateas2003facade}, sought to steer story structure toward dramatically satisfying outcomes by monitoring player actions in real time and intervening in the underlying plot. Fa\c{c}ade was a landmark demonstration: it showed that an AI system could reshape a narrative arc around player behaviour within a single, self-contained scene. But the approach was brittle in ways that limited its influence on commercial game design. Every possible intervention had to be hand-authored in advance, and the system could not generalise to player actions that fell outside the authored space. Riedl and Bulitko (\citeyear{riedl2013interactive}) surveyed the broader landscape of intelligent interactive narrative systems and diagnosed the underlying tension clearly: the more authorial control a system retains, the less it can accommodate genuine player agency, and vice versa. That tension has shaped nearly every subsequent design decision in the field.

A second wave of work approached the problem from a content generation perspective. Procedural narrative generation systems, including Tracery \citep{compton2015tracery}, the Expressionist framework \citep{kreminski2019evaluating}, and their descendants, demonstrated that grammatical expansion rules could produce textual content that varied meaningfully across play-throughs. A player entering the same tavern twice might receive different descriptions, different rumours, different atmospheric detail. Yet these systems still depended on extensive hand-authored production rules, and they had no mechanism for sensing or responding to the player's psychological state. The content varied, but it varied randomly, with no model of what the player actually needed at that moment. Short (\citeyear{short2019procedural}) crystallised the critique: procedural generation without a model of player intent risks producing content that is varied but meaningless, offering quantity of output without quality of relevance. A player who is struggling and a player who is thriving receive the same stochastic variety, which is no kind of personalization at all.

The commercial industry, meanwhile, converged on a middle path: narrative choice systems with pre-authored branches. Titles such as BioWare's \textit{Mass Effect} trilogy, CD Projekt Red's \textit{The Witcher} series, and ZA/UM's \textit{Disco Elysium} offer the player decision points whose consequences ripple through the narrative, creating the impression that the story is shaped around the individual. The impression is powerful, and at its best it produces some of the most memorable moments in the medium. But the mechanism is ultimately deterministic: the narrative delivered at any point is a fixed function of the player's explicit choice history within a pre-specified decision space. These systems do not respond to behavioural signals (play pace, combat avoidance, exploration patterns, idle time) because they were never designed to observe them. The cost of extending authored branching is a further constraint. \textit{Disco Elysium}, widely regarded as one of the most narratively ambitious RPGs ever released, contains approximately one million words of authored text, and yet it still delivers a structurally fixed narrative relative to the player's dialogue choices. Scaling this approach to accommodate implicit behavioural signals would multiply the authoring burden by orders of magnitude.

The most recent paradigm, and the one most relevant to the present work, uses generative AI to produce narrative content at runtime. PANGeA \citep{todd2024pangea} demonstrated that LLMs can generate coherent, in-world narrative when grounded in a structured knowledge base delivered through the prompt, opening the possibility of adaptation that is both infinite in variety and anchored in the game's content universe. This is a qualitative shift: the narrative space is no longer bounded by what the writer pre-authored, but by what the model can generate within a grounding constraint.

The current work situates itself at the intersection of this generative paradigm and continuous player modelling. The narrative is not merely generated but \textit{selected and styled} in response to a live, continuously updated model of the individual player's psychological state and behavioural profile, a combination that no prior system has achieved.

%--------------------------------------------------------------------------
\section{Theoretical and Conceptual Background}
%--------------------------------------------------------------------------

\subsection{Player Modelling Approaches}

\subsubsection{Bartle Taxonomy and Its Limitations}

The first systematic attempt to categorise players by what they actually want from a game, as opposed to what the game offers them, came from Bartle (\citeyear{bartle1996hearts}), who proposed four types based on his observation of Multi-User Dungeon (MUD) communities in the 1980s: Achievers, who pursue goals and progression; Explorers, who want to map the world and understand its systems; Socializers, who value interaction with other players above all else; and Killers, who derive satisfaction from dominating others. The taxonomy's enduring contribution is the foundational insight that a single game contains, in effect, several different games (one for each type of player) and that design decisions that delight one type may alienate another.

That insight remains valuable, but the taxonomy itself has well-documented weaknesses. The most fundamental is that it is categorical: each player is slotted into one of four boxes, when the empirical evidence overwhelmingly suggests that players express multiple motivations simultaneously and in shifting proportions. Yee (\citeyear{yee2006motivations}), in a large-scale factor-analytic study of MMORPG players, showed that motivation is continuous and multi-dimensional, not discrete and singular. Beyond this, the Bartle taxonomy was derived from a narrow population (players of text-based MUDs in the 1980s and 1990s) and its applicability to the visual, mechanically complex, and often single-player games of the modern era is questionable at best. The four categories are neither mutually exclusive nor operationally defined in ways that permit reliable measurement. And the principal measurement instrument, the Bartle Test, is a forced-choice self-report questionnaire that captures a one-time snapshot coloured by social desirability bias rather than a dynamic, within-session picture of what the player is actually doing.

\subsubsection{Hexad Typology}

Tondello et al.\ (\citeyear{tondello2016hexad}) set out to address these shortcomings by developing the Gamification User Types Hexad Scale, grounded explicitly in Self-Determination Theory \citep{ryan2000self}. The Hexad framework identifies six player types, each mapped to a core motivational driver: Achievers (mastery and progress), Explorers (discovery and curiosity), Socializers (social interaction and belonging), Free Spirits (autonomy and self-expression), Disruptors (challenging the status quo and testing boundaries), and Philanthropists (giving, helping, and altruism). Crucially, Hexad scores are continuous, measured on seven-point Likert items, and and a player can score highly on several dimensions at once. A player who is simultaneously a strong Explorer and a strong Philanthropist is not an anomaly under Hexad; it is the expected norm.

The original Hexad scale is a 24-item self-report questionnaire, validated for gamification contexts with acceptable internal consistency (Cronbach's $\alpha$ typically exceeding 0.70 per subscale). It has since been adopted across a range of application domains, including adaptive game design, recommendation systems, educational technology, and health interventions, establishing it as the de facto standard for research-grade player-type assessment \citep{hamari2014does}.

For the purposes of the present work, however, the Hexad scale carries a critical limitation: it is a \textit{pre-play} questionnaire. A questionnaire-derived profile is a static snapshot, captured before the player touches the game, that cannot update as the player's engagement patterns evolve over the course of a session. A player who begins as a cautious Explorer and gradually shifts toward aggressive Achiever behaviour (a common trajectory in RPGs as players gain confidence) will be mislabelled for the entire session if the profile was frozen at minute zero. Moreover, administering a 24-item questionnaire before play introduces friction that degrades ecological validity and can prime participants to think explicitly about their motivations, altering the very behaviour the system needs to observe.

This dissertation addresses the limitation head-on by inferring Hexad-analogous continuous scores directly from behavioural telemetry, updating them continuously within the session. The mapping from telemetry signals to Hexad dimensions is a novel contribution, detailed in Section~3.4.

\subsubsection{Data-Driven Player Modelling}

A more recent line of work has applied deep learning to the player modelling problem. The player2vec approach (2024) uses Transformer-based self-supervised learning to encode sequences of in-game actions into dense, low-dimensional representations, much as word2vec encodes natural language tokens into embeddings that capture semantic relationships. The result is a learned embedding space in which players with similar behavioural trajectories cluster together, without any need for hand-labelled training data or predefined type categories. Zhu et al.\ (\citeyear{zhu2023player}) surveyed the broader landscape of player modelling approaches for interactive narrative and identified the shift from questionnaire-based to behaviour-based modelling as perhaps the defining trend of the past decade.

The present work draws conceptual inspiration from player2vec, in particular the idea that a session-level embedding can capture the arc of a player's behaviour over time. But it implements a substantially simpler rule-based approach rather than a trained Transformer, for a pragmatic reason that the deep learning literature sometimes underemphasises. The cold-start problem is central: A Transformer-based session embedder needs a substantial corpus of historical play sequences (hundreds or thousands of sessions) to learn meaningful representations. For a first-session player using a newly deployed plugin with no prior data, that corpus does not exist. The rule-based Hexad profiler trades representational richness for immediate applicability: it begins producing useful profiles within the first few minutes of a session, before any learning-based method could have accumulated enough data to initialise.

More recently, Souza et al.\ (\citeyear{souza2025leveraging}) explored a complementary direction: using LLMs themselves as player modelling engines, feeding behavioural summaries into a language model and prompting it to infer player traits. The approach is attractive because it leverages the same LLM infrastructure already present in a generative narrative pipeline, but it introduces its own limitations: inference latency scales with prompt length, the model's ``understanding'' of player psychology is at best a statistical reflection of its training corpus, and the approach has not yet been validated against ground-truth psychometric instruments. The present system avoids these issues by keeping player modelling as a lightweight, deterministic computation over raw telemetry, reserving the LLM exclusively for the narrative generation task where its generative capacity is genuinely needed.

\subsection{Flow Theory in Game Design}

Csikszentmihalyi's (\citeyear{csikszentmihalyi1990flow}) theory of Flow describes a psychological state of optimal experience, the kind of complete absorption in a challenging activity that makes a person lose track of time, forget self-consciousness, and perform at the upper edge of their ability. The theory's central claim is structural: Flow arises at the intersection of high perceived challenge and high perceived skill. When challenge substantially exceeds skill, the result is anxiety: the player is overwhelmed. When skill substantially exceeds challenge, the result is boredom: the player is under-stimulated. And when both challenge and skill are low, the result is apathy: the player is disengaged in the most concerning way, neither trying nor caring. Nakamura and Csikszentmihalyi (\citeyear{nakamura2002concept}) later refined the model to emphasise that Flow is not a binary on/off state but a continuum influenced by attention, feedback loops, and the perceived meaningfulness of the activity.

The theory's influence on game design has been enormous, though often implicit rather than explicit. Chen's (\citeyear{chen2007flow}) canonical analysis of game flow was the first to make the connection formal: he argued that many of the dynamic difficulty adjustment (DDA) mechanisms already present in commercial games (adaptive enemy scaling, hint systems, rubber-banding in racing games) are, in effect, Flow-optimization mechanisms operating without the theoretical label. The challenge--skill ratio has since become the standard operationalisation of Csikszentmihalyi's balance axis in computational models of player engagement.

What has received less attention, however, is the role of \textit{narrative} in the Flow equation. DDA research has focused almost exclusively on mechanical parameters: how hard enemies hit, how fast they spawn, how generous the loot tables are. The assumption, rarely stated but pervasive, is that Flow is a function of mechanical challenge and mechanical skill. But players do not experience games as bare mechanical systems; they experience them as stories inhabited by characters in a world. A player who is mechanically in the Flow zone but narratively disengaged, receiving bland, generic descriptions while battling through a dungeon, may not report the subjective experience of Flow at all, because the narrative layer failed to sustain the sense of meaning that Flow requires.

The present work operationalises Csikszentmihalyi's model through a rule-based classifier that maps the \texttt{challenge\_skill\_ratio} (computed from behavioural telemetry) to one of four discrete states: FLOW (ratio in [0.75, 1.30]), ANXIETY (ratio $>$ 1.30), BOREDOM (ratio $<$ 0.75), and APATHY (a compound condition requiring high idle time, low dialogue engagement, \textit{and} a low challenge--skill ratio). These thresholds are informed by prior work on computational Flow detection \citep{yannakakis2009realtime} and are formalised in Section~3.4. Crucially, the reward function of the PPO adaptation engine is shaped directly around these states, making Flow the central organising principle of the entire adaptation objective: not merely a diagnostic indicator, but the thing the system is explicitly trying to produce.

%--------------------------------------------------------------------------
\section{Review of Key Methods and Technologies}
%--------------------------------------------------------------------------

\subsection{Large Language Models for Game Narrative}

\subsubsection{PANGeA (AAAI AIIDE 2024)}

The most directly relevant antecedent for the narrative generation component of this system is Todd et al.'s (\citeyear{todd2024pangea}) Procedural Artificial Narrative using Generative AI (PANGeA). The system employs an LLM to generate narrative content at runtime, grounded in a structured game knowledge base that is delivered to the model through the prompt. PANGeA's core insight is both simple and powerful: LLMs \textit{can} produce coherent, in-world narrative content, but only when the prompt supplies enough contextual grounding to anchor the model's generation within the fictional frame. Without that grounding, the model drifts: it invents characters that do not exist, references locations that contradict the map, or breaks the fourth wall by inserting contemporary idiom into a medieval setting.

PANGeA, however, stops short of personalisation. Its generation is procedural, driven by the current game state (what is happening in the world) rather than by the player's psychological state (what the player needs from the narrative at this moment). A player in Flow and a player teetering on the edge of boredom receive the same procedurally generated content, because PANGeA has no model of the player and no policy for selecting how to adapt the narrative's tone or direction. The current work extends the PANGeA paradigm by adding precisely those missing layers: a continuous player model and an RL-based adaptation policy that decides not just \textit{what} to generate but \textit{how} to modulate it.

\subsubsection{\citet{ligs2025chi}}

The LLM-Integrated Game Storytelling (LIGS) system, published at CHI 2025, approached the problem from the coherence angle. LIGS investigated how to integrate LLMs into game narrative pipelines while maintaining narrative consistency across extended play sessions, a problem that grows harder as sessions lengthen and the model must keep track of an expanding history of prior events, character states, and plot developments. LIGS introduced mechanisms for tracking narrative state across sessions and using that state to condition generation, anticipating the episodic memory approach that the current work develops independently.

A particularly valuable contribution of LIGS was its evaluation methodology. The study employed player experience questionnaires to compare AI-generated narrative against human-authored content and found that participants could not reliably distinguish the two when the LLM was appropriately constrained. This result matters for the present work: it provides empirical evidence that LLM-based narrative generation can be subjectively valid: players do not automatically perceive machine-generated text as inferior or artificial, provided the content stays within the boundaries of the game world.

\subsubsection{Knowledge-Grounded and Quest-Level Generation}

Two further lines of recent work have extended the scope of LLM-based narrative generation. Ashby et al.\ (\citeyear{ashby2023personalized}) proposed a knowledge graph-grounded approach to personalised quest and dialogue generation in RPGs, using a structured game ontology to constrain LLM output in much the same way that RAG constrains it through retrieved passages. Their system generates quest content that is consistent with the game world's entity relationships, but it does not include a real-time player model or an adaptive policy: the personalisation is structural (what quests are offered) rather than tonal (how the narrative addresses the player's current psychological state).

V{\"a}rtinen et al.\ (\citeyear{vartinen2024generating}) demonstrated that GPT-class language models can generate complete RPG quest structures, including objectives, dialogue, and reward logic, that human evaluators rated as comparable in quality to hand-authored content. This finding reinforces the viability of LLM-based narrative generation for games and complements the LIGS result on player perception. However, like PANGeA and Ashby et al., the system generates quests procedurally from game state without modelling the player or adapting the narrative tone in response to engagement signals. The present work addresses this gap by combining generation quality (via RAG grounding) with adaptation intelligence (via PPO action selection conditioned on a continuous player model).

\subsubsection{Prompt Grounding and Retrieval-Augmented Generation}

The persistent challenge of keeping LLM outputs factually and fictionally consistent with a specific game world has driven the adoption of Retrieval-Augmented Generation (RAG; \citealp{lewis2020retrieval}) in game narrative contexts. The idea is straightforward: rather than requiring the model to ``know'' the game's lore from its pre-training data (which, for a custom game world, it cannot possibly know), the system retrieves relevant passages from a curated knowledge base and injects them into the prompt at generation time. The model then generates within the semantic constraints established by the retrieved context, producing output that is grounded in the game's actual content rather than in the model's general-purpose parametric knowledge.

For game narrative, RAG offers a particularly attractive set of trade-offs. The lore corpus can be maintained, extended, and corrected independently of the model, meaning that world-building updates do not require retraining or fine-tuning. Retrieval ensures that only the passages most relevant to the current generation context are included in the prompt, avoiding the saturation of the model's finite context window. This last point is not a minor implementation detail: the Transformer architecture underlying modern LLMs \citep{vaswani2017attention} processes fixed-size context windows, and every token spent on irrelevant background is a token unavailable for the actual generation task. Efficient context usage is directly tied to generation quality.

The current system employs the sentence-transformers all-MiniLM-L6-v2 model \citep{wang2020minilm} to embed both the lore corpus and generation queries into a 384-dimensional semantic space. Retrieval operates via top-$k$ cosine similarity ($k = 3$), returning the three most relevant lore passages for inclusion in the prompt. The small model footprint (roughly 80~MB) and sub-millisecond retrieval latency make this approach well suited to real-time game narrative, where the generation pipeline must complete within the player's tolerance for loading delay.

\subsection{Reinforcement Learning for Game Adaptation}

\subsubsection{Dynamic Difficulty Adjustment}

Dynamic Difficulty Adjustment (DDA) has been an active research area since Hunicke and Chapman (\citeyear{hunicke2004ai}) demonstrated that adaptive difficulty could measurably improve player retention in \textit{Super Mario Bros}. The core idea is intuitive: if a player is dying too often, make the game easier; if the player is breezing through unchallenged, make it harder. Computational approaches to DDA have employed rule-based controllers, Bayesian models, and more recently neural networks to infer an appropriate challenge level from behavioural signals and adjust game parameters accordingly \citep{lopes2017adaptive}. The predominant formulation treats DDA as a feedback control problem: the challenge level is the controlled variable, the player's observed engagement state is the feedback signal, and the controller minimises the deviation from a target state, typically some proxy for Flow.

DDA works, and it has been deployed in commercial games ranging from racing titles (rubber-banding AI) to action games (adaptive enemy health pools). Climent et al.\ (\citeyear{climent2024framework}) recently formalised a general framework for designing RL-based DDA agents in single-player action games, demonstrating that PPO and similar policy gradient methods can learn effective difficulty policies directly from player interaction data. Their framework validates the use of RL for real-time game adaptation and provides design patterns (state representation, reward shaping, training regimes) that informed the present work's MDP formulation. But the scope of what DDA adapts remains narrow. DDA adjusts \textit{mechanical} parameters: how much damage enemies deal, how many spawn per wave, how generous the loot tables are. It does not adjust what the \textit{story does}. A player who is mechanically in the Flow zone but narratively disengaged, trudging through a dungeon with no sense of purpose because the quest text failed to land, is invisible to a DDA system.

The present work extends the DDA paradigm in two ways that, taken together, amount to a reframing. First, the adaptation target is not mechanical difficulty but \textit{narrative content}: the tone, urgency, mystery, and emotional register of the story the player receives. Second, the adaptation mechanism is a learned policy (PPO) rather than a hand-tuned feedback controller. These extensions shift the problem from mechanical tuning, which is essentially an engineering exercise, into a narrative and psychological design space where the ``right answer'' depends on who the player is and what kind of experience they are seeking.

\subsubsection{PPO and Stable-Baselines3}

Proximal Policy Optimization (PPO; \citealp{schulman2017proximal}) is a policy gradient reinforcement learning algorithm that has become the workhorse of modern RL research, owing to a combination of relative simplicity, reliable performance across diverse environments, and a clipped surrogate objective that prevents the destructively large policy updates which plagued earlier policy gradient methods. Where algorithms like TRPO imposed a hard constraint on the KL divergence between successive policies, PPO achieves a similar stabilising effect through a simpler clipping mechanism that is both easier to implement and cheaper to compute. The result is an algorithm that trains reliably with minimal hyperparameter tuning, a practical advantage that has made PPO the default choice for applications ranging from robotics to game-playing agents \citep{mnih2015human}.

Stable-Baselines3 \citep{raffin2021stable} provides production-quality PyTorch implementations of PPO and other standard RL algorithms, with interfaces compatible with the Gymnasium (formerly OpenAI Gym; \citealp{brockman2016openai}) environment specification. The library handles the engineering details (vectorised environments, gradient accumulation, logging) so that the researcher can focus on environment and reward design. The current work employs Stable-Baselines3 v2.x with the MlpPolicy architecture (two hidden layers of 64 units each), training over 200,000 timesteps in a custom Gymnasium environment that simulates narrative adaptation episodes. The use of a well-tested, reproducible RL implementation is particularly important for a system that will be evaluated on human experience: if the policy behaves unpredictably during the user study, the evaluation is compromised regardless of how well the rest of the pipeline performs.

\subsection{Episodic Memory for LLMs}

One of the most persistent limitations of Transformer-based language models is their fixed context window: the model can attend to a bounded number of tokens, and anything that falls outside that window is, for all practical purposes, forgotten. In a game narrative context, this means that an LLM generating a scene late in a play session has no access to the events it narrated twenty minutes earlier, unless those events are explicitly reintroduced into the prompt. Without some form of memory, the generated narrative will be locally coherent but globally amnesiac, referencing characters and situations that have no connection to what the player has already experienced.

Fountas et al.\ (\citeyear{fountas2024episodic}) introduced EM-LLM, an architecture that addresses this problem by endowing LLMs with a form of human-like episodic memory. The system partitions a long context stream into discrete memory events, scores each event by importance and surprise, and selectively retrieves the most relevant events for inclusion in future prompts. The result is a kind of principled compression: rather than attempting to cram the entire interaction history into a fixed-size window, the model recalls only the events most salient to the current generation context. This mirrors, at least in broad strokes, the way human episodic memory works --- not as a complete recording but as a selective reconstruction weighted toward events that were surprising, emotional, or consequential.

The current work incorporates a lightweight episodic memory inspired by EM-LLM, simplified to match the scale of the problem. The \texttt{EpisodicMemory} class maintains a capped list of up to ten narrative events, each tagged with an importance score in [0, 1] and a timestamp. When the list exceeds capacity, it is sorted by importance in descending order and truncated; the least important event is evicted, preserving the narrative moments that matter most. At generation time, the top five events by importance are retrieved and formatted as a context block in the LLM prompt, providing the model with a compressed narrative history that enables references to past events, maintains character continuity, and prevents the kind of temporal inconsistency that would break immersion.

The simplification relative to EM-LLM's full architecture (which includes surprise-based segmentation, hierarchical memory, and retrieval scoring informed by both recency and salience) is deliberate and appropriate. A 30-minute game session produces approximately 15--20 narrative events, a volume easily managed by a flat list with importance-weighted eviction. The overhead of a full hierarchical memory system would add complexity without proportional benefit at this session scale.

%--------------------------------------------------------------------------
\section{Related Work}
%--------------------------------------------------------------------------

Table~\ref{tab:related_work} draws together the key related works and maps their coverage across the six dimensions that define the scope of this dissertation. The pattern is immediately visible: each prior system addresses one or two dimensions well, but none covers more than three, and none closes the loop from player observation through adaptive generation back to in-game delivery.

\begin{table}[H]
\centering
\caption{Summary of related work on narrative personalization and player modelling.}
\label{tab:related_work}
\footnotesize
\begin{tabular}{@{}lp{3.2cm}cccccc@{}}
\toprule
\textbf{System} & \textbf{Reference} & \textbf{Player} & \textbf{Flow} & \textbf{LLM} & \textbf{RAG} & \textbf{Episodic} & \textbf{Complete} \\
 & & \textbf{Model} & \textbf{Reward} & \textbf{Gen.} & & \textbf{Memory} & \textbf{Loop} \\
\midrule
PANGeA & \citet{todd2024pangea} & No & No & Yes & Partial & No & No \\
LIGS & \citet{ligs2025chi} & No & No & Yes & No & Partial & No \\
KG-Quest & \citet{ashby2023personalized} & No & No & Yes & KG & No & No \\
GPT-Quests & \citet{vartinen2024generating} & No & No & Yes & No & No & No \\
DDA & \citet{hunicke2004ai} & Behav. & Implicit & No & No & No & No \\
RL-DDA & \citet{climent2024framework} & Behav. & Implicit & No & No & No & No \\
LLM-PM & \citet{souza2025leveraging} & LLM & No & No & No & No & No \\
player2vec & \citet{devlin2024player2vec} & Deep & No & No & No & No & No \\
EM-LLM & \citet{fountas2024episodic} & No & No & Yes & No & Yes & No \\
\textbf{This work} & & \textbf{Hexad+} & \textbf{Explicit} & \textbf{Yes} & \textbf{Yes} & \textbf{Yes} & \textbf{Yes} \\
 & & \textbf{Flow} & & \textbf{(RAG)} & & & \\
\bottomrule
\end{tabular}
\end{table}

The gap is not that these individual techniques are unknown; each has been demonstrated in isolation. The gap is that no one has wired them together. Continuous Hexad profiling derived from behavioural telemetry, Flow-based RL reward shaping, RAG-grounded LLM generation, and episodic session memory have never been integrated into a single closed-loop system that runs in real time within a game engine. The present work addresses that gap by combining all six components into a deployable Unity plugin and evaluating the result through a mixed-methods study that measures both the psychological impact (via the GEQ and miniPXI) and the subjective perception (via the NPS-3 and semi-structured interviews).
