%##########################################################################
% CHAPTER 4: SYSTEM IMPLEMENTATION
%##########################################################################
\chapter{System Implementation}

\section{Chapter Overview}

This chapter describes how the five-module system specified in Chapter~3 was actually built. The gap between a methodology section and working code is often wider than either party admits; this chapter bridges it by documenting the technology choices, architectural patterns, and implementation details that turned the design into a running system. It covers the technology stack, the backend service, the database layer, each of the three pipeline stages (player modelling, narrative generation, RL adaptation), the Unity plugin, and the testing procedures used to validate correctness.

\section{Technology Stack}

The system is implemented across two platforms: a Python backend service and a Unity C\# plugin. Table~\ref{tab:backend_stack} and Table~\ref{tab:unity_stack} summarize the key dependencies.

\begin{table}[H]
\centering
\caption{Backend technology stack.}
\label{tab:backend_stack}
\small
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Component} & \textbf{Technology} & \textbf{Version} \\
\midrule
Runtime & Python & 3.14 \\
API framework & FastAPI & 0.115+ \\
Data validation & Pydantic & v2 \\
Database & SQLite & (stdlib) \\
RL framework & Stable-Baselines3 & 2.x \\
RL environment & Gymnasium & 0.29+ \\
Embeddings & sentence-transformers & latest \\
LLM client & httpx (Ollama API) & 0.27+ \\
Numerical & NumPy & 2.x \\
Statistics & SciPy, pingouin & latest \\
Data analysis & pandas & 2.x \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Unity plugin and inference stack.}
\label{tab:unity_stack}
\small
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Component} & \textbf{Technology} & \textbf{Version} \\
\midrule
Engine & Unity & 6 LTS \\
Language & C\# & .NET 9 \\
HTTP client & UnityWebRequest & built-in \\
JSON & Newtonsoft.Json & 13.x \\
UI & TextMeshPro & built-in \\
\midrule
LLM runtime & Ollama (local) & --- \\
Primary model & Phi-3.5 Mini & 3.8B params \\
Fallback model & Llama 3.2 3B & --- \\
\bottomrule
\end{tabular}
\end{table}

\section{Backend Service}

The backend is a single FastAPI application that serves as the computational core of the system. It receives telemetry from the Unity client, runs the player modelling pipeline, selects narrative actions via the PPO agent, generates narrative content through the LLM, and returns the result --- all within a single request--response cycle that must complete before the next telemetry batch arrives five seconds later.

The application exposes four primary endpoints:

\begin{itemize}
\item \texttt{POST /api/telemetry} --- accepts a \texttt{TelemetryBatch} payload, persists it to SQLite, triggers player modelling, and initiates the adaptation cycle.
\item \texttt{POST /api/narrative/generate} --- accepts a session identifier and returns the generated \texttt{NarrativeContent}, blocking until generation is complete.
\item \texttt{GET /api/player-model/\{session\_id\}} --- returns the current \texttt{PlayerModel} including Hexad profile, Flow state, and session metadata.
\item \texttt{WebSocket /ws/telemetry} --- low-latency telemetry streaming channel.
\end{itemize}

The server is configured with CORS middleware to accept connections from the Unity client's localhost origin. All data is persisted to SQLite with separate tables for telemetry batches, player models, and narrative events. Pydantic v2 models handle all input validation and output serialization, ensuring type safety at the API boundary.

The asynchronous architecture of FastAPI allows the narrative generation pipeline --- including the 8-second Ollama timeout --- to execute without blocking other incoming telemetry requests. This is critical for maintaining the 5-second telemetry cycle even when LLM generation is in progress.

\section{Database Layer}

The SQLite schema uses three primary tables:

\begin{enumerate}
\item \textbf{telemetry\_batches} --- stores all 23 \texttt{TelemetryBatch} fields plus a server-side \texttt{received\_at} timestamp. Indexed on \texttt{(player\_id, session\_id, window\_start)}.
\item \textbf{player\_models} --- stores \texttt{PlayerModel} snapshots indexed on \texttt{(player\_id, session\_id, timestamp)}. Each telemetry POST triggers an upsert of the computed model.
\item \textbf{narrative\_events} --- stores \texttt{NarrativeContent} records (action, content\_type, content, speaker, emotional\_tone, fallback flag) with foreign keys to session\_id. This log enables post-hoc analysis of which narrative actions were fired at what points in each session.
\end{enumerate}

SQLite was selected over PostgreSQL for three reasons that all pointed in the same direction. First, simplicity of deployment: SQLite requires no external service, no configuration, and no connection pooling --- the database is a single file on disk that the application creates on first run. Second, the study scale is modest: $N = 50$ participants, each generating roughly 360 telemetry batches over a 30-minute session, produces a total volume that SQLite handles without breaking a sweat. Third, the access pattern is single-writer: only one participant plays at a time during the controlled study, so the write concurrency limitations of SQLite are irrelevant.

\section{Player Modelling Implementation}

The player modelling pipeline is implemented as three sequential classes called within the same request handler:

\begin{lstlisting}[language=Python, caption={Player modelling pipeline flow.}]
all_batches = TelemetryBatch[] (full session from SQLite)
recent_batches = all_batches[-6:]  # ~30s sliding window

recent_batches
  -> FeatureExtractor.extract()  -> dict[str, float] (11 features)
  -> FlowClassifier.classify()   -> (FlowState, float) (state + conf)

all_batches
  -> HexadProfiler.compute()     -> HexadProfile      (6 dims)

  -> PlayerModel (assembled)
\end{lstlisting}

The modelling service uses a split-window strategy: the \texttt{FeatureExtractor} receives the most recent six \texttt{TelemetryBatch} records (approximately 30~seconds), producing features that reflect the player's current engagement state. The \texttt{FlowClassifier} operates on these recent features to classify the instantaneous Flow state. In contrast, the \texttt{HexadProfiler} receives all session batches, because player-type signals emerge cumulatively over minutes rather than seconds. This split ensures that Flow classification is responsive to momentary changes while Hexad profiling benefits from the full behavioural history.

Figure~\ref{fig:hexad_radar} shows an illustrative Hexad profile for a hypothetical player at the 15-minute mark of a session, demonstrating how the six dimensions can form distinct behavioural signatures.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=1.1]
% ── Radar chart: 6 axes ──

% ── Grid rings ──
\foreach \r in {0.2, 0.4, 0.6, 0.8, 1.0} {
    \draw[black!15, line width=0.3pt]
        (90:\r*2.8) -- (30:\r*2.8) -- (-30:\r*2.8) --
        (-90:\r*2.8) -- (-150:\r*2.8) -- (150:\r*2.8) -- cycle;
    \node[font=\tiny, text=black!35] at (90:\r*2.8+0.22) {\pgfmathprintnumber{\r}};
}

% ── Axis lines and labels ──
\foreach \ang/\lbl in {90/Achiever, 30/Explorer, -30/Socializer,
                       -90/{Free Spirit}, -150/Disruptor, 150/Philanthropist} {
    \draw[black!25, line width=0.3pt] (0,0) -- (\ang:2.8);
    \node[font=\scriptsize, align=center] at (\ang:3.3) {\lbl};
}

% ── Player A: Explorer-Philanthropist (blue) ──
\draw[blue!60, line width=1.2pt, fill=blue!10, fill opacity=0.3]
    (90:0.45*2.8) -- (30:0.85*2.8) -- (-30:0.55*2.8) --
    (-90:0.70*2.8) -- (-150:0.20*2.8) -- (150:0.80*2.8) -- cycle;
\foreach \ang/\val in {90/0.45, 30/0.85, -30/0.55, -90/0.70, -150/0.20, 150/0.80} {
    \fill[blue!70] (\ang:\val*2.8) circle (1.5pt);
}

% ── Player B: Achiever-Disruptor (red) ──
\draw[red!60, line width=1.2pt, fill=red!10, fill opacity=0.3]
    (90:0.90*2.8) -- (30:0.30*2.8) -- (-30:0.25*2.8) --
    (-90:0.35*2.8) -- (-150:0.85*2.8) -- (150:0.20*2.8) -- cycle;
\foreach \ang/\val in {90/0.90, 30/0.30, -30/0.25, -90/0.35, -150/0.85, 150/0.20} {
    \fill[red!70] (\ang:\val*2.8) circle (1.5pt);
}

% ── Legend ──
\draw[blue!60, line width=1.2pt] (1.5,-3.5) -- ++(0.6,0);
\node[font=\scriptsize, right] at (2.2,-3.5) {Player A (Explorer / Philanthropist)};
\draw[red!60, line width=1.2pt] (1.5,-3.9) -- ++(0.6,0);
\node[font=\scriptsize, right] at (2.2,-3.9) {Player B (Achiever / Disruptor)};

\end{tikzpicture}
\caption{Illustrative Hexad radar profiles for two hypothetical players at the 15-minute mark. Player~A exhibits strong Explorer and Philanthropist signals (high lore engagement, high exploration coverage). Player~B exhibits strong Achiever and Disruptor signals (high kill count, high objective completion). These contrasting profiles would trigger different narrative actions from the PPO agent.}
\label{fig:hexad_radar}
\end{figure}

A notable implementation detail --- small but critical --- is the denominator protection pattern applied throughout: \texttt{max(deaths, 1)}, \texttt{max(obj\_attempted, 1)}, and so on for every ratio computation. In the opening minutes of a session, many counters are still zero: the player has not died, has not attempted an objective, has not found a lore item. Without these guards, the feature extractor would raise a division-by-zero exception on the very first telemetry batch, crashing the pipeline before the player had done anything at all. The \texttt{max(x, 1)} pattern is simple, defensive, and costs nothing; it ensures that the system produces sensible (if uninformative) feature values from the moment the session begins.

\section{Narrative Pipeline Implementation}

The narrative pipeline is invoked asynchronously following each player model update. The process follows four stages:

\begin{enumerate}
\item \textbf{Action selection:} The \texttt{NarrativeAgent.select\_action()} method constructs the 7-dimensional observation vector from the \texttt{PlayerModel} and calls \texttt{model.predict(obs, deterministic=False)}.
\item \textbf{RAG retrieval:} The query string is constructed from the Flow state, current activity, and location to maximize retrieval relevance. The top-3 lore chunks are returned.
\item \textbf{Prompt assembly:} The \texttt{PromptBuilder} combines the system prompt, player context, retrieved lore, episodic memory context, and action-specific instructions.
\item \textbf{LLM generation:} The \texttt{OllamaClient} sends the prompt and returns the parsed \texttt{NarrativeContent}, with fallback handling for malformed or timed-out responses.
\end{enumerate}

Stochastic prediction (\texttt{deterministic=False}) is used at inference time to maintain behavioural diversity. With deterministic prediction, the agent would always select the single highest-probability action for a given state, which in practice means the player would receive the same narrative modulation every time they entered a particular Flow state. Stochastic sampling introduces controlled variation: the agent still favours the best action, but occasionally selects alternatives, producing a more varied and naturalistic narrative experience.

The importance score of 0.7 assigned to new memory events reflects a baseline estimate --- a reasonable default in the absence of a more nuanced importance model. A more sophisticated implementation could derive importance from the narrative action type: LORE\_REWARD events, which represent significant story revelations, would receive higher importance than NO\_CHANGE events, which represent deliberate narrative silence. This refinement is left for future work.

\section{RL Adaptation Implementation}

The \texttt{NarrativeAdaptationEnv} is registered as a standard Gymnasium environment and trained using Stable-Baselines3 PPO. Table~\ref{tab:ppo_hyperparams} lists the hyperparameters.

\begin{table}[H]
\centering
\caption{PPO hyperparameters.}
\label{tab:ppo_hyperparams}
\small
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Hyperparameter} & \textbf{Value} \\
\midrule
Policy & MlpPolicy (2 $\times$ 64 hidden layers) \\
Learning rate & $3 \times 10^{-4}$ \\
n\_steps & 512 \\
batch\_size & 64 \\
n\_epochs & 10 \\
gamma ($\gamma$) & 0.95 \\
gae\_lambda & 0.95 \\
clip\_range & 0.2 \\
n\_envs (vectorized) & 4 \\
total\_timesteps & 200,000 \\
\bottomrule
\end{tabular}
\end{table}

Training uses four parallel vectorised environments (\texttt{make\_vec\_env}) to improve sample efficiency: each environment runs an independent episode, and the four experience streams are interleaved to reduce correlation between consecutive training samples. The trained model is saved to \texttt{./data/ppo\_narrative\_agent.zip} and loaded on subsequent server starts, so the training cost is paid once and the inference cost is negligible. Episode length is set to 360 steps --- one step per 5-second telemetry window, corresponding to 30 minutes of simulated gameplay --- matching the expected duration of a user study session.

Figure~\ref{fig:ppo_convergence} shows the representative training convergence curve. The agent's mean episodic reward rises from near-zero (random policy) to a plateau around $+250$ over the 200,000 training timesteps, indicating successful learning of Flow-promoting actions.

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.88\textwidth,
    height=6.5cm,
    xlabel={Training Timesteps ($\times 10^3$)},
    ylabel={Mean Episodic Reward},
    xmin=0, xmax=200,
    ymin=-50, ymax=320,
    grid=major,
    grid style={black!8},
    axis lines=left,
    tick label style={font=\small},
    label style={font=\small},
    legend style={font=\scriptsize, at={(0.02,0.98)}, anchor=north west, draw=none, fill=white, fill opacity=0.8},
    smooth,
    thick,
]

% ── Mean reward curve (representative) ──
\addplot[blue!70, line width=1.2pt] coordinates {
    (0, -15) (5, -8) (10, 12) (15, 35) (20, 58)
    (25, 72) (30, 95) (35, 110) (40, 128) (45, 140)
    (50, 155) (55, 162) (60, 170) (65, 178) (70, 185)
    (75, 192) (80, 198) (85, 205) (90, 210) (95, 218)
    (100, 222) (105, 228) (110, 232) (115, 236) (120, 238)
    (125, 241) (130, 244) (135, 246) (140, 248) (145, 249)
    (150, 250) (155, 252) (160, 253) (165, 254) (170, 254)
    (175, 255) (180, 256) (185, 256) (190, 257) (195, 257) (200, 258)
};
\addlegendentry{Mean episodic reward}

% ── Variance band (approximate) ──
\addplot[blue!15, line width=0pt, fill=blue!10, forget plot] coordinates {
    (0, -40) (10, -15) (20, 30) (40, 95) (60, 140) (80, 168)
    (100, 192) (120, 210) (140, 222) (160, 230) (180, 235) (200, 238)
} -- (200, 278) --
    plot coordinates {
    (200, 278) (180, 277) (160, 276) (140, 274) (120, 266)
    (100, 252) (80, 228) (60, 200) (40, 161) (20, 86)
    (10, 39) (0, 10)
} -- cycle;

% ── Annotation: convergence region ──
\draw[black!40, dashed, line width=0.5pt] (axis cs:120,0) -- (axis cs:120,280);
\node[font=\tiny, text=black!50, rotate=90, anchor=south] at (axis cs:122,140) {Policy stabilises};

\end{axis}
\end{tikzpicture}
\caption{PPO training convergence. Mean episodic reward (blue line) with approximate variance band (shaded) over 200,000 timesteps. The agent transitions from near-random behaviour to a stable policy that maintains the simulated player in FLOW for the majority of each episode. The dashed line marks the approximate convergence point ($\sim$120k steps).}
\label{fig:ppo_convergence}
\end{figure}

\section{Unity Plugin Implementation}

The Unity plugin is structured as a UPM-compatible package. The \texttt{PlayerDataLogger} accumulates telemetry via direct field increments in \texttt{Update()} and fires the HTTP POST at fixed 5-second intervals using a Coroutine.

The \texttt{NarrativeManager} uses a polling architecture: immediately following each telemetry POST's response, it fires a GET request to \texttt{/narrative/generate}. The 8-second Ollama generation timeout is handled by displaying no new content rather than blocking the game loop. The \texttt{ContentInjector} implements a state machine with three states: IDLE, DISPLAYING, and COOLING\_DOWN. Content queued during COOLING\_DOWN is held and injected after the 3-second minimum spacing, ensuring narrative delivery does not overwhelm the player during intense combat sequences.

For the demo game integration, additional scripts bridge the game engine's event system to the plugin's telemetry API: \texttt{TelemetryHooks.cs} connects combat, quest, and dialogue events to the corresponding \texttt{PlayerDataLogger} recording methods, while \texttt{ExplorationTracker.cs} implements a grid-based tile exploration counter that updates the \texttt{tiles\_explored} field. \texttt{NarrativeUI.cs} provides the TMPro-based HUD panel with auto-dismiss and fade animations.

\section{Testing and Validation}

Testing a system that spans three technology stacks (Python, C\#, Ollama) and five conceptual modules requires a layered approach. The testing strategy proceeds from fine-grained unit tests through integration tests to system-level validation.

\textbf{Unit tests} were written using pytest, targeting the components where incorrect behaviour would propagate most damagingly through the pipeline. The \texttt{FeatureExtractor} tests cover edge cases that arise naturally in early-session telemetry: empty batch lists, zero-denominator fields (the player has not died, has not attempted an objective), and maximum value clamping (a player who achieves a kill--death ratio well above the normalisation ceiling). The \texttt{HexadProfiler} tests verify that initialisation with empty batches returns the expected 0.5 neutral defaults rather than undefined or zero values. The \texttt{FlowClassifier} tests probe boundary conditions at the critical ratio thresholds of 0.75 and 1.30, as well as the APATHY compound condition (which requires three simultaneous criteria). The \texttt{compute\_reward} tests exercise all state transitions, confirm that the streak bonus triggers after three consecutive FLOW steps, and verify that the repetition penalty activates when three identical actions are selected in succession.

\textbf{Integration tests} validated the end-to-end pipeline from a mock \texttt{TelemetryBatch} POST request through player modelling, action selection, narrative generation, and JSON response. These tests confirmed two critical properties: that the Flow state correctly influences the tone directive in the LLM prompt (an ANXIETY state produces a ``supportive'' tone, not an ``exciting'' one), and that the fallback mechanism activates cleanly when Ollama is unavailable or returns malformed output.

\textbf{Validation of the simulated MDP} was conducted by examining the trained PPO agent's convergence behaviour. The agent's mean episodic reward rose substantially above zero over the 200,000 training timesteps, indicating that it had learned to prefer helpful actions over random or degenerate ones. More informatively, the agent's action distribution in ANXIETY states was dominated by PROVIDE\_GUIDANCE and LOWER\_STAKES --- exactly the actions the system designer intended for overwhelmed players --- while BOREDOM states elicited INCREASE\_URGENCY and ADD\_MYSTERY. These distributions provide confidence that the reward function's incentive structure is translating into the intended policy behaviour.

\textbf{Lore retrieval validation} was conducted by manually querying the retriever with ten representative in-session queries (e.g., ``combat in the contested zone,'' ``the chronicler's history of the vale'') and verifying that retrieved chunks were semantically relevant to the query. All ten queries returned at least one clearly relevant chunk in the top-3 results, and seven returned two or more. This confirms that the all-MiniLM-L6-v2 model provides adequate semantic discrimination for a corpus of this size and specificity.
